---
title: "Visualizing Amounts"
output:
  html_document:
    df_print: paged
    keep_md: true
date: "2024-02-14"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
library(dplyr)
library(ggplot2)
library(gapminder)
library(lattice)

```

# 1. Dot plots

One of the simplest visualizations of a single numerical variable with a modest number of observations and labels for the observations is a dot plot, or Cleveland dot plot:

```{r}
# Filtering data
le_am_2007 <- filter(gapminder, 
                     year == 2007,
                     continent == "Americas")
# Setting up the theme
thm <- theme_minimal() +
  theme(text = element_text(size = 16))

# Visualize
ggplot(le_am_2007, aes(x = lifeExp,
                       y = country)) +
  geom_point(color = "deepskyblue3", size = 2) +
  labs(x = "Life Expectancy (years)",
       y = NULL) + 
  thm # Apply the theme

```

This visualization:

-   Shows the overall distribution of the data

-   Makes it easy to locate the life expectancy of a particular country.

#### Unless there is a natural order to the categories (e.g. months of the year or days of the week) it is usually better to `reorder` to make the plot increasing or decreasing:

```{r}
ggplot(le_am_2007, aes(x = lifeExp,
                       y = reorder(country, lifeExp))) +
  geom_point(color = "deepskyblue3",
             size = 2) +
  labs(x = "Life Expectancy (years)",
       y = NULL) +
  thm

```

-   Locating a particular country is a little more difficult.

-   But the shape of the distribution is more apparent.

-   Approximate median and quartiles can be read off easily.

#### Dot plot uses:

Dot plots are particularly appropriate for `interval` data.

-   They often do not show the origin;

-   They focus the viewer's attention on differences.

Dot plots are often very useful for **group summaries** like *totals* or *averages*.

For the barley data, total yield within each site, adding up across all varieties and both years, can be computed as:

```{r}
b_tot_site <- group_by(barley, site) |>
  summarize(yield = sum(yield))

ggplot(b_tot_site, aes(x = yield,
                       y = site)) +
  geom_point(color = "deepskyblue3",
             size = 2.5) +
  labs(x = "Total Yield (bushels/acre)",
       y = NULL) +
  thm

```

#### Larger Data Sets

For larger data sets, like the `citytemps` data with 140 observations, over-plotting of labels becomes a problem:

```{r}
citytemps <- read.table("http://www.stat.uiowa.edu/~luke/data/citytemps.dat", header = TRUE)
ggplot(citytemps, aes(x = temp,
                      y = reorder(city, temp))) +
  geom_point(color = "deepskyblue3",
             size = 0.5) +
  labs(x = "Temperature (degrees F)",
       y = NULL) +
  thm

```

Reducing to 30 or 40, e.g. by taking a sample or a meaningful subset, can help:

```{r}
ct1 <- filter(citytemps, temp < 32) |> sample_n(10)
ct2 <- filter(citytemps, temp >= 32) |> sample_n(20)
ctsamp <- bind_rows(ct1, ct2)

ggplot(ctsamp, aes(x = temp,
                   y = reorder(city, temp))) +
  geom_point(color = "deepskyblue3",
             size = 2) +
  labs(x = "Temperature (degrees F)",
       y = NULL) +
  thm
```

#### Some Variations

The size of the dots can be used to encode an additional numeric variable.

This view uses area to encode population size:

This is sometimes called a **bubble chart**.

```{r}
ggplot(le_am_2007, aes(x = lifeExp,
                       y = reorder(country, lifeExp),
                       size = pop / 1000000)) +
  geom_point(color = "deepskyblue3") +
  labs(x = "Life Expectancy (years)",
       y = NULL) +
  scale_size_area("Population\n(Millions)", 
                  max_size = 8) +
  thm + theme(legend.position = "top")

```


# 2. Visualizing Three or More Numeric Variables

## 2.1. Scatterplot Matrices
A scatterplot matrix is a useful overview that shows all pairwise scatterplots.

There are many options for creating scatterplot matrices in R; a few are:

-   `pairs` in base graphics;

-   `splom` in package `lattice`

-   `ggpairs` in `GGally`.

Some examples using the `mpg` data:

```{r}
library(lattice)
library(ggplot2)
library(dplyr)

splom(select(mpg, cty, hwy, displ),
        cex = 0.5, pch = 19)


```
```{r}
library(GGally)

ggpairs(select(mpg, cty, hwy, displ),
        lower = list(continuous = 
                       wrap("points",
                            size = 1)))


```
### 2.1.1 Three Data Sets

Thee useful data sets to explore:

The `ethanol` data frame in the `lattice` package.

Soil resistivity data from from Cleveland's Visualizing Data book.

The `quakes` data frame in the `datasets` package.

#### 2.1.1.1 Ethanol Data

The `ethanol` data frame in the `lattice` package contains data from an experiment on efficiency and emissions in small one-cylinder engines.

The data frame contains 88 observations on three variables:

`NOx`: Concentration of nitrogen oxides (NO and NO2) in micrograms.

`C` Compression ratio of the engine.

`E` Equivalence ratio, a measure of the richness of the air and ethanol fuel mixture.

```{r}
library(lattice)

splom(select(ethanol, NOx, C, E),
      cex = 0.5, pch = 19)


```
A goal is to understand the relationship between the pollutant `NOx` and the controllable variables `E` and `C`.

#### 2.1.1.2 Earth Quake Locations and Magnitudes
The `quakes` data frame contains data on locations of seismic events of magnitude 4.0 or larger in a region near Fiji.

The time frame is from 1964 to perhaps 2000.

More recent data is available from a number of sources on the web.

A scatter plot matrix:

```{r}
library(lattice)
splom(quakes, cex = 0.1, pch = 19)

```
Quake locations:

Some goals:

-   understand the three-dimensional location of the quakes;

-   see if there is any association between location and magnitude.

```{r, warning=FALSE}
library(maps)
library(mapproj)

md <- map_data("world2", c("Fiji", "Tonga", "New Zealand"))

ggplot(quakes, aes(x = long,
                   y = lat)) +
  geom_polygon(aes(group = group), data = md, color = "black", fill = NA) +
  geom_point(size = 0.5, color = "red")+
  coord_map() +
  ggthemes::theme_map()


```

```{r}



```























